Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic

### 2024-11-25 manuscript/manuscript.qmd
A bunch of proofreading copyedits

Our findings make two main contributions to the literatures on human rights, international law, and democratic backsliding. First, we find that international legal mechanisms such as derogations did provide some protection against human rights abuses during the pandemic, even in states experiencing democratic backsliding. This finding speaks to scholarship on international legal compliance [@ConradRitter:2019; @Hillebrecht:2014], showing that treaty actions that allow for temporary suspension of legal obligations do not always contribute to more violations and abuse. Second, our results show that democratic backsliding states are still interested in appearing as legitimate members of the international community, take actions to protect their reputations, and project transparency to international audiences during times of crises. These results have important implications for how the international community may be able to prevent further autocratization in backsliding states.

@@

Combating a deadly virus like COVID-19 required extraordinary public health measures that often conflicted with personal rights and freedoms. Globally, IOs were confronted with how to situate the crises within global governance and human rights frameworks [@Comstock:2024], as were national level governments [@ChaudhryComstockHeiss:2024]. One way that states navigated the pandemic was to enact emergency measures. Internationally, states could commit to these emergency provisions via treaty derogations. Treaty derogations, or temporary suspensions of states' international treaty commitments, are intended to allow flexibility to states while they are experiencing a crisis---which could be civil conflict, natural disaster, or a public health crisis. Derogations provide vital information to international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows actors---at least in principle---to challenge measures that are excessive, vague, or outlast the in- tended time frame of their implementation [@Helfer:2021]. The International Covenant on Civil and Political Rights (ICCPR) is the only treaty with universal UN membership that requires states to issue derogations when human rights are suspended during times of crises, making it the main focus of our empirical analysis. The ICCPR places clear limitations on state actions when derogating, including that they must be proportionate, non-discriminatory, and temporary. It also specifies certain rights as non-derogable and outlines a formal procedure for derogation: the emergency must be formally declared and the UN Secretary General must be notified.

Extant research on derogations has examined the kinds of states that are more likely to utilize them and the conditions under which they do so [@Burchill:2005; @HBHelferFariss]. This research finds that democracies are more likely to issue derogations and other post-commitment actions related to human rights treaties [@Comstock:2019; @Simmons:2009]. Though the most frequent derogators are stable democracies and countries where domestic courts can exercise strong oversight of the executive and hold them responsible for breaches of human rights agreements, derogations---especially in the context of a long-lasting event such like a pandemic [@Helfer:2021; @Lebret:2020]---are more concerning because of the tendency of some countries to turn into "serial derogators." These countries generally derogate "without providing information about rights restrictions and in multiple consecutive years" [@HBHelferFariss, p. 675]. Other research has also found that governments frequently violate both derogable and non-derogable rights [@richardsclay2012]. This paper builds on research by examining the dynamics between democratic backsliding and state behavior as it relates to international human rights law derogations. Extending analysis to democratic backsliders allows us to understand how regimes experiencing democratic erosion or authoritarian resurgence might engage with these international legal actions.

Global democratic backsliding was already in motion when the pandemic spread in March 2020 [@LuhrmannLindberg:2020; @MaerzLuhrmannHellmeier:2020]. As of 2019, the share of countries experiencing democratic erosion more than doubled in the past decade compared to the decade before [@IDEA:2019], with only 8% of the world's population living in countries becoming more democratic [@LuhrmannLindberg:2020]. Democratic backsliding is a particular subset of such erosion involving the intentional weakening of checks and balances, and the curtailment and rollbacks of civil liberties, political rights, personal freedoms, and a broad swath of human rights. @fig-panback-map demonstrates that North America and Western Europe generally had low risk of democratic backsliding during the pandemic. However, this does not indicate that there was no backsliding occurring. Researchers point to election manipulation and executive overreach in the United States as examples of democratic erosion [@Williamson:2023] and the European Union has responded via sanctions to curb democratic backsliding of member states [@BlaubergerSedelmeier:2024]. The higher risk areas included parts of South America, across Africa, India, and some parts of Eastern Europe and Southeast Asia. Overall, pandemic-era democratic backsliding was a global phenomenon. @fig-derogation-map depicts states that issued pandemic-related derogations to the ICCPR. Most states do not issue derogations even during times of emergency, so the determinants of when they are issued is important to understand. Geographically, derogations were issued across different regions including South America, Africa, Eastern Europe, and some parts of Asia. They were not generally issued across North America or Western Europe.

@@

Existing research on regime transitions, human rights compliance, and commitment are informative for our expectations about backsliding states and legal engagement. Regime transition matters along several important fronts. Regimes transitioning toward democracy, or democratizing, are more likely to use the international human rights regime as a means of legitimation [@Moravcsik:2000], including through treaty commitment [@Comstock:2021]. Though new democracies are more likely to commit to broad human rights conventions more quickly, they are also more hesitant to commit to human rights treaties with more demands [@DaiTokhi:2023]. Regimes transitioning away from democracy---or experiencing democratic backsliding---are found to repress human rights more. @Davenport:1999, for example, finds that the transition towards autocratization has a significant and positive impact on repressive behavior while democratization promotes more open governance. The strength of domestic democratic institutions, in particular, is important for these findings [@HillJones:2014]. @Meyerrose:2020 argues that if IOs promote some domestic institutions but not others, backsliding can occur when domestic executives are empowered above other institutions. In examining democratic backsliding and human rights more specifically, @AdhikariKingMurdie:2024 find strong statistical support that democratic backsliding harmed human rights. The authors found that the degree or intensity of backsliding along with the length of the backsliding period contributed to further chances of human rights violations. @Ginsburg:2019 suggests that international and regional courts may be a means for activists to resist the spread of national-level democratic backsliding. @MechkovaLuhrmannLindberg:2017 caution against interpreting democratic backsliding as too extensive of a trend, noting that most liberal and electoral democracies that declined were unlikely "to collapse so fully that it becomes an autocracy" (165). Rather, democratic character remains, at least in part, in these states experiencing democratic backsliding wherein the "normative power of democracy remains relatively strong" (168).

@@

Bringing these literatures together, we expect that states that issue derogations are more willing and capable of communicating to domestic and international audiences about potential and real rights restrictions during the COVID-19 pandemic. Concerns about international reputation often shape behavior and are connected to the desire to obtain certain goals---these usually include some form of aid, trade, membership and holding office in IOs, and receiving higher rankings on performance indicators [@BushZetterberg:2021; @KelleySimmons:2019; @LevitskyWay:2010]. Generally, when countries seek cooperation from the international community, they seek external legitimacy [@Kelley:2012] and will consequently abide by legal and normative commitments to uphold human rights and rule of law.

Public international commitments, such as compliance with the ICCPR, can thus be an important way for states, especially those backsliding, to signal their intent to the international community, to agree to respect these rights moving forward. Human rights compliance can be a costly signal for the costs associated with both reaching human rights standards and the threat of penalty if the standards are not met [@Moore:2003]. Human rights compliance and legal engagement provides information about states to others in the international community. More powerful or prominent states may feel the need to signal less given the information readily available about their interests and behavior [@Moore:2003, 890]. However, states with less information available about their interests, commitment, and behavior can use human rights treaty signaling to convey that information to a broad, global audience.

@@

To test these expectations, we conduct a series of quantitative analyses followed by a brief examination of illustrative cases mapping on to each quadrant of @tbl-expectations (we include more detailed case studies in the appendix). From these series of tests, we provide a deeper understanding of motivations and behavior of states during times of a health crisis and global threats to democracy.

@@

We use weekly data from March 11, 2020 to June 30, 2021 for 139 countries during the first 69 weeks of the pandemic. Table A1 in the appendix lists the main variables we used across our models. To measure democratic backsliding, we use the Pandemic Backsliding Index (PanBack) from the Varieties of Democracy (V-Dem) project [@vdem-v12; @vdem-pandem]. PanBack measures the risk of democratic backsliding specifically during the pandemic, assessing how state responses to the pandemic violated democratic standards, based on the severity of different kinds of human rights violations. PanBack ranges from 0 to 1, with high values representing greater risk of backsliding. Following V-Dem, we dichotomize values and consider PanBack scores above 0.3 as high risk.

@@

To explore whether democratic backsliding had an overall impact on legal behavior toward international human rights treaties, @fig-treaty-action-state-type descriptively shows the count of all human rights treaty actions by states, divided by type of filing. Derogations are overwhelming the most common type of international treaty action issued during the pandemic. In the appendix, we run a model that predicts non-derogation actions (such ratifications, reservations, and declarations) using the same covariates used to predict ICCPR derogations, and find that the rule of law index is the only factor that significantly predicts non-derogation human rights treaty actions---backsliding is not substantially associated with other kinds of treaty actions. Examining these treaty actions as a dependent variable allows us to test whether democratic backsliding states participated with international law, broadly, in different ways from their counterparts and/or had a unique set of behaviors with more critical legal actions such as derogations.These findings together may indicate that for general, non-emergency legal actions, strong domestic rule of law and institutions generally matter for international legal behavior, but during times of crisis, other factors are at play. Backsliding behavior did not shape states' legal engagement with all human rights treaties---rather, states specifically and uniquely responded to the pandemic with targeted derogations.

@@

The predicted probabilities for public transportation measures and stay-at-home restrictions demonstrate similar patterns (@fig-restrictions-preds, panels B and C). Non-backsliding, non-derogating states are the most likely to have no emergency measures (`r preds_pubtrans_l$low_backsliding_risk.no_derogation.no_measures$prob_ci_nice` and `r preds_stayhome_l$low_backsliding_risk.no_derogation.no_measures$prob_ci_nice`, respectively) and least likely to impose the harshest restrictions (`r preds_pubtrans_l$low_backsliding_risk.no_derogation.require_closing$prob_ci_nice` and `r preds_stayhome_l$low_backsliding_risk.no_derogation.require_not_leaving_house_with_minimal_exceptions$prob_ci_nice` respectively). For non-backsliding states, having a derogation in place is associated with a substantially higher probability of requiring the closing of transportation systems (`r preds_pubtrans_l$low_backsliding_risk.derogation.require_closing$prob_ci_nice`; twice as likely) and requiring residents to stay home (`r preds_stayhome_l$low_backsliding_risk.derogation.require_not_leaving_house_with_minimal_exceptions$prob_ci_nice`; nearly two-thirds as likely). States with a high risk of backsliding that do not derogate appear roughly similar to non-backsliding, derogating states, while high-risk states that derogated are the most likely to require closing (`r preds_pubtrans_l$high_backsliding_risk.derogation.require_closing$prob_ci_nice`) and the least likely to have no emergency measures at all (`r preds_pubtrans_l$high_backsliding_risk.derogation.no_measures$prob_ci_nice`). Derogations appear particularly important for stay-at-home measures, which are the most individualized of the three emergency measures here (i.e. residents are constrained to their homes vs. prohibited from traveling between cities or using public transportation). More democratic, non-backsliding states appear hesitant to impose requirements to stay-at-home, and the states that did so reinforced this emergency violation with a derogation. States at greater risk of backsliding were more willing to impose these restrictions without derogations, and those that derogated had the highest probability of strict emergency measures.

@@

Derogation status also matters for the risk of not having time-limited emergency measures. For backsliding states, derogation is associated with a `r diffs_hr_nolimit_l$high_backsliding_risk.derogation_no_derogation.none$pp_ci_nice` percentage point higher probability of having time limited measures (`r diffs_hr_nolimit_l$high_backsliding_risk.derogation_no_derogation.none$p_d`), while in non-backsliding states there is a `r diffs_hr_nolimit_l$low_backsliding_risk.derogation_no_derogation.none$pp_ci_nice` percentage point difference, which is larger and more precise (`r diffs_hr_nolimit_l$low_backsliding_risk.derogation_no_derogation.none$p_d`). This difference can be directly attributable to derogations, since the derogation process requires that states specify time limits for their emergency measures.

@@

We briefly consider four states as illustrative cases of varying derogation and democratic backsliding behavior to better unpack how democratic backsliding and derogation interaction (or lack thereof) may have influenced state behavior during the pandemic. Earlier in the paper we posited that transparency about restrictions and violation would vary across states based on derogation and backsliding behavior. @tbl-mini-cases situates the cases across backsliding and derogation status. We selected cases that depict both regional variation and low/high risk of backsliding. Including extreme cases should be an "easy test" of the influence of backsliding on human rights violations and other behaviors if democratic backsliding has an impact. The four cases illustrate derogations' mitigating effect in states experiencing democratic backsliding and those that did not. Derogating states in both backsliding scenarios had concerns about maintaining international reputation and legitimacy (Armenia, Guatemala), and derogations helped them communicate their pandemic measures in a more or less transparent manner. However, states not as interested in protecting their international reputation, especially in the perspective of Western states, were less likely to use derogations as a means to communicate their intentions about pandemic measures (India, Hungary). These states subsequently engaged in more abusive and discriminatory enforcement of such measures. The dynamics illustrated in these cases overall support the purpose of derogations in international law. However, state motivations for declaring derogations among states both experiencing backsliding or not are more complex and merits further investigation. A full discussion of each of the cases is included in the appendix.

@@

The statistical findings indicate a few key takeaways. First, derogation behavior was distinct from other types of human rights treaty legal behaviors during the pandemic. States that did not typically submit treaty actions did so intentionally during the pandemic to signal the restriction of human rights obligations to the ICCPR. Second, backsliding states still engaged in signaling crisis restrictions on rights via the submission of derogations. Democratic backsliders did not withdraw from participating with the international human rights regime, they still saw some value in participating. Third, backsliding was associated with increased odds of abusive and discriminatory practices during the pandemic. Fourth, and importantly, the interaction between derogations and backsliding appears to have potentially mitigated some of the abusive behaviors during the pandemic.

What these findings point to is that democratic backsliding is a complex phenomenon---states experiencing backsliding impose discriminatory policies and abusive behaviors, but part of their institutional embeddedness remains invested, at least in part, in communicating some of these problems and maintaining connections with the international community around human rights. It is important to note that this study maps behavior that occurred but it does not fully theorize or test why backsliders were motivated to care about and submit derogations. We encourage future research on democratic backsliding and human rights to more fully examine backslider consideration of reputation, foreign policy during backsliding, and specifically international human rights law behavior during backsliding. Research might also track how continued engagement with international human rights law might shape, or mitigate, backsliding itself. Future research also would benefit the study of whether and how derogations might be used after a crisis has ended. Most states did not formally derogate, but nearly all states implemented similar emergency policies. States that did not rely on derogations implemented emergency policies by stretching domestic norms and laws to accommodate expanded government policies. Do states continue using crisis justification after a local/regional/global crisis is largely over, and does the formality of derogation influence the probability of returning to pre-crisis policies? Newer research hints that derogations were associated with a reduced probability of human rights abuses early in the pandemic and that states that filed derogations did not use temporary measures as a carte blanche "pandemic pass" for additional repression [@ChaudhryComstockHeiss:2024]. However, it is too early to tell if formal emergency declarations will encourage or prevent a return to pre-pandemic policies. Overall, our findings contribute to the understanding of how neither derogation filing nor backsliding alone depict a complete picture of human rights and legal behaviors during the pandemic.

>>>>

Our findings make two main contributions to the literature on human rights, international law, and democratic backsliding. First, we find that international legal mechanisms such as derogations did provide some protection against human rights abuses during the pandemic, even in states experiencing democratic backsliding. This finding speaks to scholarship on international legal compliance [@ConradRitter:2019; @Hillebrecht:2014], showing that treaty actions that allow for temporary suspension of legal obligations do not always contribute to more violations and abuse. Second, our results show that democratic backsliding states are still interested in appearing as legitimate members of the international community, take actions to protect their reputations, and project transparency to international audiences during times of crises. These results have important implications for how the international community may be able to prevent further autocratization in backsliding states.

@@

Combating a deadly virus like COVID-19 required extraordinary public health measures that often conflicted with personal rights and freedoms. Globally, IOs were confronted with how to situate the crises within global governance and human rights frameworks [@Comstock:2024], as were national-level governments [@ChaudhryComstockHeiss:2024]. One way that states navigated the pandemic was to enact emergency measures. Internationally, states could commit to these emergency provisions via treaty derogations. Treaty derogations, or temporary suspensions of states' international treaty commitments, are intended to allow flexibility to states while they are experiencing a crisis---which could be civil conflict, natural disaster, or a public health crisis. Derogations provide vital information to international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows actors---at least in principle---to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@Helfer:2021]. The International Covenant on Civil and Political Rights (ICCPR) is the only treaty with universal UN membership that requires states to issue derogations when human rights are suspended during times of crises, making it the main focus of our empirical analysis. The ICCPR places clear limitations on state actions when derogating, including that they must be proportionate, non-discriminatory, and temporary. It also specifies certain rights as non-derogable and outlines a formal procedure for derogation: the emergency must be formally declared, and the UN Secretary-General must be notified.

Extant research on derogations has examined the kinds of states that are more likely to utilize them and the conditions under which they do so [@Burchill:2005; @HBHelferFariss]. This research finds that democracies are more likely to issue derogations and other post-commitment actions related to human rights treaties [@Comstock:2019; @Simmons:2009]. Though the most frequent derogators are stable democracies and countries where domestic courts can exercise strong oversight of the executive and hold them responsible for breaches of human rights agreements, derogations---especially in the context of a long-lasting event like a pandemic [@Helfer:2021; @Lebret:2020]---are more concerning because of the tendency of some countries to turn into "serial derogators." These countries generally derogate "without providing information about rights restrictions and in multiple consecutive years" [@HBHelferFariss, p. 675]. Other research has also found that governments frequently violate both derogable and non-derogable rights [@richardsclay2012]. This paper builds on research by examining the dynamics between democratic backsliding and state behavior as it relates to international human rights law derogations. Extending analysis to democratic backsliders allows us to understand how regimes experiencing democratic erosion or authoritarian resurgence might engage with these international legal actions.

Global democratic backsliding was already in motion when the pandemic spread in March 2020 [@LuhrmannLindberg:2020; @MaerzLuhrmannHellmeier:2020]. As of 2019, the share of countries experiencing democratic erosion more than doubled in the past decade compared to the decade before [@IDEA:2019], with only 8% of the world's population living in countries becoming more democratic [@LuhrmannLindberg:2020]. Democratic backsliding is a particular subset of such erosion involving the intentional weakening of checks and balances, and the curtailment and rollbacks of civil liberties, political rights, personal freedoms, and a broad swath of human rights. @fig-panback-map demonstrates that North America and Western Europe generally had low risk of democratic backsliding during the pandemic. However, this does not indicate that there was no backsliding occurring. Researchers point to election manipulation and executive overreach in the United States as examples of democratic erosion [@Williamson:2023] and the European Union has responded via sanctions to curb democratic backsliding of member states [@BlaubergerSedelmeier:2024]. The higher risk areas included parts of South America, across Africa, India, and some parts of Eastern Europe and Southeast Asia. Overall, pandemic-era democratic backsliding was a global phenomenon. @fig-derogation-map depicts states that issued pandemic-related derogations to the ICCPR. Most states do not issue derogations even during times of emergency, so the determinants of when they are issued are important to understand. Geographically, derogations were issued across different regions including South America, Africa, Eastern Europe, and some parts of Asia. They were not generally issued across North America or Western Europe.

@@

Existing research on regime transitions, human rights compliance, and commitment is informative for our expectations about backsliding states and legal engagement. Regime transition matters along several important fronts. Regimes transitioning toward democracy, or democratizing, are more likely to use the international human rights regime as a means of legitimation [@Moravcsik:2000], including through treaty commitment [@Comstock:2021]. Though new democracies are more likely to commit to broad human rights conventions more quickly, they are also more hesitant to commit to human rights treaties with more demands [@DaiTokhi:2023]. Regimes transitioning away from democracy---or experiencing democratic backsliding---are found to repress human rights more. @Davenport:1999, for example, finds that the transition towards autocratization has a significant and positive impact on repressive behavior while democratization promotes more open governance. The strength of domestic democratic institutions, in particular, is important for these findings [@HillJones:2014]. @Meyerrose:2020 argues that if IOs promote some domestic institutions but not others, backsliding can occur when domestic executives are empowered above other institutions. In examining democratic backsliding and human rights more specifically, @AdhikariKingMurdie:2024 find strong statistical support that democratic backsliding harmed human rights. The authors found that the degree or intensity of backsliding along with the length of the backsliding period contributed to further chances of human rights violations. @Ginsburg:2019 suggests that international and regional courts may be a means for activists to resist the spread of national-level democratic backsliding. @MechkovaLuhrmannLindberg:2017 caution against interpreting democratic backsliding as too extensive of a trend, noting that most liberal and electoral democracies that declined were unlikely "to collapse so fully that it becomes an autocracy" (165). Rather, democratic character remains, at least in part, in these states experiencing democratic backsliding wherein the "normative power of democracy remains relatively strong" (168).

@@

Bringing these literatures together, we expect that states that issue derogations are more willing and capable of communicating to domestic and international audiences about potential and real rights restrictions during the COVID-19 pandemic. Concerns about international reputation often shape behavior and are connected to the desire to obtain certain goals---these usually include some form of aid, trade, membership and holding office in IOs, and receiving higher rankings on performance indicators [@BushZetterberg:2021; @KelleySimmons:2019; @LevitskyWay:2010]. Generally, when countries seek cooperation from the international community, they seek external legitimacy [@Kelley:2012] and will consequently abide by legal and normative commitments to uphold human rights and the rule of law.

Public international commitments, such as compliance with the ICCPR, can thus be an important way for states, especially those backsliding, to signal their intent to the international community, to agree to respect these rights moving forward. Human rights compliance can be a costly signal for the costs associated with both reaching human rights standards and the threat of penalty if the standards are not met [@Moore:2003]. Human rights compliance and legal engagement provide information about states to others in the international community. More powerful or prominent states may feel the need to signal less given the information readily available about their interests and behavior [@Moore:2003, 890]. However, states with less information available about their interests, commitment, and behavior can use human rights treaty signaling to convey that information to a broad, global audience.

@@

To test these expectations, we conduct a series of quantitative analyses followed by a brief examination of illustrative cases mapping onto each quadrant of @tbl-expectations (we include more detailed case studies in the appendix). From these series of tests, we provide a deeper understanding of motivations and behavior of states during times of a health crisis and global threats to democracy.

@@

We use weekly data from March 11, 2020, to June 30, 2021, for 139 countries during the first 69 weeks of the pandemic. Table A1 in the appendix lists the main variables we used across our models. To measure democratic backsliding, we use the Pandemic Backsliding Index (PanBack) from the Varieties of Democracy (V-Dem) project [@vdem-v12; @vdem-pandem]. PanBack measures the risk of democratic backsliding specifically during the pandemic, assessing how state responses to the pandemic violated democratic standards, based on the severity of different kinds of human rights violations. PanBack ranges from 0 to 1, with high values representing greater risk of backsliding. Following V-Dem, we dichotomize values and consider PanBack scores above 0.3 as high risk.

@@

To explore whether democratic backsliding had an overall impact on legal behavior toward international human rights treaties, @fig-treaty-action-state-type descriptively shows the count of all human rights treaty actions by states, divided by type of filing. Derogations are overwhelmingly the most common type of international treaty action issued during the pandemic. In the appendix, we run a model that predicts non-derogation actions (such as ratifications, reservations, and declarations) using the same covariates used to predict ICCPR derogations, and find that the rule of law index is the only factor that significantly predicts non-derogation human rights treaty actions---backsliding is not substantially associated with other kinds of treaty actions. Examining these treaty actions as a dependent variable allows us to test whether democratic backsliding states participated with international law, broadly, in different ways from their counterparts and/or had a unique set of behaviors with more critical legal actions such as derogations. These findings together may indicate that for general, non-emergency legal actions, strong domestic rule of law and institutions generally matter for international legal behavior, but during times of crisis, other factors are at play. Backsliding behavior did not shape states' legal engagement with all human rights treaties---rather, states specifically and uniquely responded to the pandemic with targeted derogations.

@@

The predicted probabilities for public transportation measures and stay-at-home restrictions demonstrate similar patterns (@fig-restrictions-preds, panels B and C). Non-backsliding, non-derogating states are the most likely to have no emergency measures (`r preds_pubtrans_l$low_backsliding_risk.no_derogation.no_measures$prob_ci_nice` and `r preds_stayhome_l$low_backsliding_risk.no_derogation.no_measures$prob_ci_nice`, respectively) and least likely to impose the harshest restrictions (`r preds_pubtrans_l$low_backsliding_risk.no_derogation.require_closing$prob_ci_nice` and `r preds_stayhome_l$low_backsliding_risk.no_derogation.require_not_leaving_house_with_minimal_exceptions$prob_ci_nice`, respectively). For non-backsliding states, having a derogation in place is associated with a substantially higher probability of requiring the closing of transportation systems (`r preds_pubtrans_l$low_backsliding_risk.derogation.require_closing$prob_ci_nice`; twice as likely) and requiring residents to stay home (`r preds_stayhome_l$low_backsliding_risk.derogation.require_not_leaving_house_with_minimal_exceptions$prob_ci_nice`; nearly two-thirds as likely). States with a high risk of backsliding that do not derogate appear roughly similar to non-backsliding, derogating states, while high-risk states that derogated are the most likely to require closing (`r preds_pubtrans_l$high_backsliding_risk.derogation.require_closing$prob_ci_nice`) and the least likely to have no emergency measures at all (`r preds_pubtrans_l$high_backsliding_risk.derogation.no_measures$prob_ci_nice`). Derogations appear particularly important for stay-at-home measures, which are the most individualized of the three emergency measures here (i.e. residents are constrained to their homes vs. prohibited from traveling between cities or using public transportation). More democratic, non-backsliding states appear hesitant to impose requirements to stay-at-home, and the states that did so reinforced this emergency violation with a derogation. States at greater risk of backsliding were more willing to impose these restrictions without derogations, and those that derogated had the highest probability of strict emergency measures.

@@

Derogation status also matters for the risk of not having time-limited emergency measures. For backsliding states, derogation is associated with a `r diffs_hr_nolimit_l$high_backsliding_risk.derogation_no_derogation.none$pp_ci_nice` percentage point higher probability of having time-limited measures (`r diffs_hr_nolimit_l$high_backsliding_risk.derogation_no_derogation.none$p_d`), while in non-backsliding states there is a `r diffs_hr_nolimit_l$low_backsliding_risk.derogation_no_derogation.none$pp_ci_nice` percentage point difference, which is larger and more precise (`r diffs_hr_nolimit_l$low_backsliding_risk.derogation_no_derogation.none$p_d`). This difference can be directly attributable to derogations, since the derogation process requires that states specify time limits for their emergency measures.

@@

We briefly consider four states as illustrative cases of varying derogation and democratic backsliding behavior to better unpack how democratic backsliding and derogation interaction (or lack thereof) may have influenced state behavior during the pandemic. Earlier in the paper, we posited that transparency about restrictions and violations would vary across states based on derogation and backsliding behavior. @tbl-mini-cases situates the cases across backsliding and derogation status. We selected cases that depict both regional variation and low/high risk of backsliding. Including extreme cases should be an "easy test" of the influence of backsliding on human rights violations and other behaviors if democratic backsliding has an impact. The four cases illustrate derogations' mitigating effect in states experiencing democratic backsliding and those that did not. Derogating states in both backsliding scenarios had concerns about maintaining international reputation and legitimacy (Armenia, Guatemala), and derogations helped them communicate their pandemic measures in a more or less transparent manner. However, states not as interested in protecting their international reputation, especially in the perspective of Western states, were less likely to use derogations as a means to communicate their intentions about pandemic measures (India, Hungary). These states subsequently engaged in more abusive and discriminatory enforcement of such measures. The dynamics illustrated in these cases overall support the purpose of derogations in international law. However, state motivations for declaring derogations among states both experiencing backsliding or not are more complex and merit further investigation. A full discussion of each of the cases is included in the appendix.

@@

The statistical findings indicate a few key takeaways. First, derogation behavior was distinct from other types of human rights treaty legal behaviors during the pandemic. States that did not typically submit treaty actions did so intentionally during the pandemic to signal the restriction of human rights obligations to the ICCPR. Second, backsliding states still engaged in signaling crisis restrictions on rights via the submission of derogations. Democratic backsliders did not withdraw from participating with the international human rights regime---they still saw some value in participating. Third, backsliding was associated with increased odds of abusive and discriminatory practices during the pandemic. Fourth, and importantly, the interaction between derogations and backsliding appears to have potentially mitigated some of the abusive behaviors during the pandemic.

What these findings point to is that democratic backsliding is a complex phenomenon---states experiencing backsliding impose discriminatory policies and abusive behaviors, but part of their institutional embeddedness remains invested, at least in part, in communicating some of these problems and maintaining connections with the international community around human rights. It is important to note that this study maps behavior that occurred, but it does not fully theorize or test why backsliders were motivated to care about and submit derogations. We encourage future research on democratic backsliding and human rights to more fully examine backslider consideration of reputation, foreign policy during backsliding, and specifically international human rights law behavior during backsliding. Research might also track how continued engagement with international human rights law might shape or mitigate backsliding itself. Future research also would benefit the study of whether and how derogations might be used after a crisis has ended. Most states did not formally derogate, but nearly all states implemented similar emergency policies. States that did not rely on derogations implemented emergency policies by stretching domestic norms and laws to accommodate expanded government policies. Do states continue using crisis justification after a local/regional/global crisis is largely over, and does the formality of derogation influence the probability of returning to pre-crisis policies? Newer research hints that derogations were associated with a reduced probability of human rights abuses early in the pandemic and that states that filed derogations did not use temporary measures as a carte blanche "pandemic pass" for additional repression [@ChaudhryComstockHeiss:2024]. However, it is too early to tell if formal emergency declarations will encourage or prevent a return to pre-pandemic policies. Overall, our findings contribute to the understanding of how neither derogation filing nor backsliding alone depict a complete picture of human rights and legal behaviors during the pandemic.
